---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---

**<span style="color:red">Note: You will need the following packages: dplyr & timeDate </span>**

## Loading and preprocessing the data
```{r}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url=fileurl,destfile="./activity.zip",method="curl")
unzip("./activity.zip")
activity <- read.csv("activity.csv")

#Here is what it looks like
head(activity,10)
```



## What is mean total number of steps taken per day?
**1. Calculate the total number of steps taken per day**
```{r}
library(dplyr)

tidy_data <- activity %>%  filter(!is.na(steps)) %>% group_by(date) %>% summarise_each(funs(sum))
#Here is what it looks like
head(tidy_data,10)
```


**2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day**
```{r}
#Histogram
hist(tidy_data$steps, main = "Histogram of Steps per day", xlab = "Steps", col=rgb(0,1,1,0.4), ylim=c(0,35))

#Here is what a barplot looks like
barplot(tidy_data$steps, names.arg = c(as.character(tidy_data$date)), cex.axis=0.5, cex.names = 0.5 , las=2 , main = "Barplot of Steps per day")
```


**3. Calculate and report the mean and median of the total number of steps taken per day**
```{r}
tidy_data <- activity %>%  filter(!is.na(steps)) %>% group_by(date) %>% summarise_each(funs(sum,mean,median)) %>% select(date,steps_sum,steps_mean,steps_median)
#Here is what it looks like
head(tidy_data,10)
```



## What is the average daily activity pattern?
**1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**
```{r}
tidy_data1 <- activity %>%  filter(!is.na(steps) ) %>% select(interval,steps) %>% group_by(interval) %>% summarise_each(funs(mean))

plot(tidy_data1$interval,tidy_data1$steps,type = "l",ylab = "Average across dates",xlab = "Interval", cex.axis=0.5)
```


**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r}
tidy_data1  %>%  filter( steps == max(steps))
```



## Imputing missing values
**1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**
```{r}
activity %>%  filter(is.na(steps) ) %>% nrow()
```


**2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

<i><span style="color:blue">The strategy I have used is to replace the 'NA' with the mean steps for that particualr interval across all days. Here is how its implemented in code:</span></i>
```{r}
chkaa <- merge(activity, tidy_data1, by.x = "interval", by.y = "interval")
names(chkaa)[2]<-"steps"
names(chkaa)[4]<-"meansteps"
names(chkaa)[3]<- "date"
```


**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**
```{r}
chkaa1 <- chkaa %>% mutate(steps = ifelse(is.na(steps),meansteps,steps)) %>% select(date,steps,interval)
#Here is what it looks like
head(chkaa1,10)
```


**4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**
```{r}
chkaa2 <- chkaa1 %>% group_by(date) %>% summarise_each(funs(sum,mean,median)) %>% select(date,steps_sum,steps_mean,steps_median)

#Plotting graphs for comparison side-by-side
par(mfcol = c(1,2))
hist(chkaa2$steps_sum, main = "AFTER substituting for NA", ylim=c(0,35),xlab = "Steps", col=rgb(1,1,0,0.7))
hist(tidy_data$steps_sum, ylim=c(0,35),  col=rgb(0,1,1,0.4), main = "BEFORE substituting for NA", xlab = "Steps")

#Plotting graphs for comparison by overlaying on top of each other
par(mfcol = c(1,1))
hist(chkaa2$steps_sum, main = "Overlaying Histogram of Steps per day -- BEFORE & AFTER", ylim=c(0,35),xlab = "Steps", col=rgb(1,1,0,0.7))
par(new=TRUE)
hist(tidy_data$steps_sum, ylim=c(0,35),axes = FALSE, labels = FALSE, col=rgb(0,1,1,0.4), main = NULL, xlab = NULL)
legend("topright", c("BEFORE","AFTER","OVERLAP"), lty = c(1,1,1) , lwd=10, col=c(rgb(0,1,1,0.4),rgb(1,1,0,0.7),rgb(0,1,0,0.4)), bty = "n" )

```

<i><span style="color:blue">The frequency of steps between 10000 - 15000 only increased, rest of it is exactly similar.</span></i>

```{r}
#Here is a what the sum, mean, median data looks like:
head(chkaa2,10)

```



## Are there differences in activity patterns between weekdays and weekends?
**1. Create a new factor variable in the dataset with two levels "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**
```{r}
library(timeDate)
wkday <- chkaa1 %>% mutate(daydate = ifelse(isWeekday(as.Date(chkaa1$date)),"Weekday","Weekend")) %>% select(date,daydate,steps,interval)
#Here is what it looks like
head(wkday,10)
```


**2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.**
```{r}
library(lattice)
chkaa3 <- wkday %>% select(daydate,interval,steps) %>% group_by(daydate,interval) %>% summarise_each(funs(mean))
xyplot(chkaa3$steps ~ chkaa3$interval| chkaa3$daydate , type = "l",ylab = "Average Number of steps",xlab = "Interval", cex.axis=0.5 , layout= c(1,2))
```

